Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-05-08 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
update
trying to decrease the intro and combine it with viscou fingering

Numerical investigation of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterisation of such fluid flows serves as the foundation of reservoir engineering and groundwater studies \citep{white_1981}. More recently, due to the role of green house gases (GHG) emissions in the global climate change, several research work have focused on CO$_{2}$ migration and trapping mechanisms in carbon capture utilisation and storage (CCUS) operations \citep{spycher_2003, self_2012, jiang_2011}.

\medskip
Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities,][]{saffman_1959}. This work focuses on flow instabilities in two-phase systems due to viscous and stress forces by neglecting gravitational effects (often referred as viscous instabilities or viscous fingering). %Section \ref{section:ViscousInstabilities}).

\medskip
In oil and gas reservoir exploration, viscous and density instabilities are relatively common during water-injection processes. As water and oil interacts, the interface between these fluids moves creating an uneven or fingered flow profile (see Fig.~\ref{fig:simple_case}). Viscous fingering results in inefficient flow sweeping which can bypass significant quantities of recoverable oil and may lead to early breakthrough of water into neighbour production wells. This problem is associated with water-flooding processes and caused by high-permeability layers and unfavorable mobility ratios. Water breakthrough and volumetric sweep efficiency (ratio of volumes between the recovered and the injected fluids) are the main determinants of the productive life of a reservoir \citep{riaz_2004, tavassoli_2015}. During immiscible CO$_{2}$-flooding $\left(\text{\ie in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, viscosity of supercritical CO$_{2}$ is lower than crude oil, thus viscous fingering and/or channelling may occur.

\medskip
Viscous instabilities can be found across several disciplines and scales and are mainly controlled by the mobility ratio (MR) between displacing and displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (\ie wide spatial porosity and/or permeability distribution), gravitational forces, anisotropic dispersion, non-monotonic viscosity profile etc \citep{budek_2017,nicolaides_2015}. \citet{muskat_1934} investigated fluid flow in Hele-Shaw cells, \ie parallel flat plates separated by an infinitesimal gap, and the impact on the capillary number ($N_{c}$),

>>>>

Numerical investigation of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterisation of such fluid flows serves as the foundation of reservoir engineering and groundwater studies \citep{white_1981}. %More recently, due to the role of green house gases (GHG) emissions in the global climate change, several research work have focused on CO$_{2}$ migration and trapping mechanisms in carbon capture utilisation and storage (CCUS) operations \citep{spycher_2003, self_2012, jiang_2011}.
Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities,][]{saffman_1959}. This work focuses on flow instabilities in two-phase systems due to viscous and stress forces by neglecting gravitational effects (often referred as viscous instabilities or viscous fingering). %Section \ref{section:ViscousInstabilities}).

\medskip
In oil and gas reservoir exploration, viscous and density instabilities are relatively common during water-injection processes. As water and oil interacts, the interface between these fluids moves creating an uneven or fingered flow profile (see Fig.~\ref{fig:simple_case}). Viscous fingering results in inefficient flow sweeping which can bypass significant quantities of recoverable oil and may lead to early breakthrough of water into neighbour production wells. This problem associated with water-flooding processes is caused by high-permeability layers and unfavorable mobility ratios. Water breakthrough and volumetric sweep efficiency (ratio of volumes between the recovered and the injected fluids) are the main determinants of the productive life of a reservoir \citep{riaz_2004, tavassoli_2015}. During immiscible CO$_{2}$-flooding $\left(\text{\ie in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, viscosity of supercritical CO$_{2}$ is lower than crude oil, thus viscous fingering and/or channelling may occur.

\medskip
Viscous instabilities can be found across several disciplines and scales and are mainly controlled by the mobility ratio (MR)(see eq.\ref{eq:MR}) between displacing and displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (\ie wide spatial porosity and/or permeability distribution), gravitational forces, anisotropic dispersion, non-monotonic viscosity profile etc \citep{budek_2017,nicolaides_2015}. \citet{muskat_1934} investigated fluid flow in Hele-Shaw cells, \ie parallel flat plates separated by an infinitesimal gap, and the impact on the capillary number ($N_{c}$),
